---
title: "DATA 607 - Week 5 Assignment"
author: "Brian K. Liles"
date: "March 3, 2018"
output: html_document
---

#Tidying and Transforming Data
Import the flights csv file which describes arrivals and delays for two airlines across five destinations. The CSV file was created and exported in MySQL Workbench 6.3 CE.

#Load libraries
```{r}
library(tidyverse)
library(tidyr)
library(dplyr)
library(knitr)
```
#Read the flights csv file
```{r}
flights <- read_csv("https://raw.githubusercontent.com/LilesB/DATA-607-5WK-Assignment/master/flights.csv")
glimpse(flights)
```
The **flights** tibble has 4 observations and 8 variables. In order to create a clean environment, we will remove the AID primary key variable and then use the **kable** function to display the contents of the tibble.
```{r}
flights <- subset(flights,select=-AID)
kable(flights)
```
Use the **gather()** function to condense the original 4 observations & 7 variable tibble to a tibble with 20 observations & 4 variables and create a tibble entitled **flights_gather**
```{r}
flights_gather <- flights %>% gather(Destination,NumFlights,-Airline,-FlightStatus)
kable(flights_gather)
```
Next, we use the **spread()** function to condense **flights_long** to a tibble with 10 observations & 4 variables named **flights_spread**

```{r}
flights_spread <- flights_gather %>% spread(FlightStatus,NumFlights)
kable(flights_spread)
```
#Analyze the data
In order to analyze the data first the **glimpse** function was used in order to see if the **delayed** and **on time** variables were integers. 

After utilizing the **spread()** function we see that **on time** needs to be renamed which was done immediately after the **glimpse** function.
```{r}
glimpse(flights_spread)
```

```{r}
colnames (flights_spread) <- c("Airline", "Destination", "Delayed","OnTime")
names(flights_spread)
```
Next we create a variable entitled **TotFlights** which adds the **Delayed** and **OnTime** variables.
```{r}
flights_spread <- flights_spread %>% mutate(TotFlights = Delayed + OnTime)
flights_spread %>% select(Airline,Destination,TotFlights)
```
Next we create a variable entitled **DelayedRatio** which divides the **Delayed** variable by **TotFlights**
```{r}
flights_spread <- flights_spread %>% mutate(DelayedRatio = Delayed/TotFlights) 
flights_spread %>% select(Airline,Destination,DelayedRatio)
```
Next we create a variable entitled **OnTimeRatio** which divides the **OnTime** variable by **TotFlights**
```{r}
flights_spread <- flights_spread %>% mutate(OnTimeRatio = OnTime/TotFlights) 
flights_spread %>% select(Airline,Destination,OnTimeRatio)
```
#Analysis

##Top Five Airline/Destination combo - Delayed Flights
```{r}
LowDel5 <- arrange(flights_spread, Delayed)
LowDel5 %>% select(Airline:Delayed) %>% slice(1:5)
```
##Bottom Five Airline/Destination combo - Delayed Flights
```{r}
TopDel5 <- arrange(flights_spread,desc (Delayed))
TopDel5 %>% select(Airline:Delayed) %>% slice(1:5)
```
##Bottom Five Airline/Destination combo - On Time Flights
```{r}
LowOnTime5 <- arrange(flights_spread, OnTime)
LowOnTime5 %>% select(Airline,Destination,OnTime) %>%
slice(1:5)
```
##Top Five Airline/Destination combo - On Time Flights
```{r}
TopOnTime5 <- arrange(flights_spread, desc(OnTime))
TopOnTime5 %>% select(Airline,Destination,OnTime) %>%
slice(1:5)
```
##Bottom Five Airline/Destination combo - Total Number of Flight
```{r}
LowTotFlights5 <- arrange(flights_spread, TotFlights)
LowTotFlights5 %>% select(Airline,Destination,TotFlights) %>%
slice(1:5)
```
##Top Five Airline/Destination combo - Total Number of Flight
```{r}
TopTotFlights5 <- arrange(flights_spread, desc(TotFlights))
TopTotFlights5 %>% select(Airline,Destination,TotFlights) %>%
slice(1:5)
```
##Top Five Airline/Destination combo - Delayed Ratio
```{r}
LowDelRats5 <- arrange(flights_spread, DelayedRatio)
LowDelRats5 %>% select(Airline,Destination,DelayedRatio) %>%
slice(1:5)
```
##Bottom Five Airline/Destination combo - Delayed Ratio
```{r}
TopDelRats5 <- arrange(flights_spread, desc(DelayedRatio))
TopDelRats5 %>% select(Airline,Destination,DelayedRatio) %>%
slice(1:5)
```
##Bottom Five Airline/Destination combo - On Time Ratio
```{r}
LowOnTimeRat5 <- arrange(flights_spread, OnTimeRatio)
LowOnTimeRat5 %>% select(Airline,Destination,OnTimeRatio) %>%
slice(1:5)
```
##Top Five Airline/Destination combo - On Time Ratio
```{r}
TopOnTimeRat5 <- arrange(flights_spread, desc(OnTimeRatio))
TopOnTimeRat5 %>% select(Airline,Destination,OnTimeRatio) %>%
slice(1:5)
```
##Delayed Data Element
```{r}
delayed <- flights_spread %>% group_by(Airline) %>% summarise(DelayedMean=mean(Delayed), DelayedSD=sd(Delayed), DelayedMed=median(Delayed))
delayed
```
```{r}
boxplot(flights_spread$Delayed ~ flights_spread$Airline, main = "Delayed Flights Box Plot ALASKA / AM WEST Airlines", xlab = "Airlines", ylab = "Number of Delayed Flights", col = c("yellow","beige"))
```
```{r}
barplot(delayed$DelayedMean, horiz=FALSE, col = c("yellow","beige"),main="ALASKA / AM West: Delayed Flights Mean", names.arg=c("ALASKA","AM WEST"), xlab = "Airlines")
```
##On Time Data Element
```{r}
on_time <- flights_spread %>% group_by(Airline) %>% summarise(OnTimeMean=mean(OnTime), OnTimeSD=sd(OnTime), OnTimeMed=median(OnTime))
on_time
```
```{r}
boxplot(flights_spread$OnTime ~ flights_spread$Airline, main = "On Time Flights Box Plot ALASKA / AM WEST Airlines", xlab = "Airlines", ylab = "Number of On Time Flights", col = c("yellow","beige"))
```
```{r}
barplot(on_time$OnTimeMed, horiz=FALSE, col = c("yellow","beige"),main="ALASKA / AM West: On Time Flights Median", names.arg=c("ALASKA","AM WEST"), xlab = "Airlines")
```
##Total Flights Data Element
```{r}
total_flights <- flights_spread %>% group_by(Airline) %>% summarise(TotFlightsMean=mean(TotFlights), TotFlightsSD=sd(TotFlights), TotFlightsMed=median(TotFlights))
total_flights
```
```{r}
boxplot(flights_spread$TotFlights ~ flights_spread$Airline, main = "Total Flights Box Plot ALASKA / AM WEST Airlines", xlab = "Airlines", ylab = "Number of Flights", col = c("yellow","beige"))
```
```{r}
barplot(total_flights$TotFlightsMed, horiz=FALSE, col = c("yellow","beige"),main="ALASKA / AM West: Total Number of  Flights Median", names.arg=c("ALASKA","AM WEST"), xlab = "Airlines")
```
##Delayed Ratio Data Element
```{r}
delayed_ratio <- flights_spread %>% group_by(Airline) %>% summarise(DelRatMean=mean(DelayedRatio), DelRatSD=sd(DelayedRatio), DelRatMed=median(DelayedRatio))
delayed_ratio
```
```{r}
boxplot(flights_spread$DelayedRatio ~ flights_spread$Airline, main = "Delayed Ratio Box Plot ALASKA / AM WEST Airlines", xlab = "Airlines", ylab = "Number of Flights", col = c("yellow","beige"))
```
```{r}
round(delayed_ratio$DelRatMean,digits = 2)
barplot(delayed_ratio$DelRatMean, horiz=FALSE, col = c("yellow","beige"),main="ALASKA / AM West: Delayed Ratio", names.arg=c("ALASKA","AM WEST"), xlab = "Airlines")
```
##On Time Ratio Data Element
```{r}
ontime_ratio <- flights_spread %>% group_by(Airline) %>% summarise(OnTRatMean=mean(OnTimeRatio), OnTRatSD=sd(OnTimeRatio), OnTRatMed=median(OnTimeRatio))
ontime_ratio
```
```{r}
boxplot(flights_spread$OnTimeRatio ~ flights_spread$Airline, main = "On Time Ratio Box Plot ALASKA / AM WEST Airlines", xlab = "Airlines", ylab = "Number of Flights", col = c("yellow","beige"))
```
```{r}
round(ontime_ratio$OnTRatMed,digits = 2)
barplot(ontime_ratio$OnTRatMed, horiz=FALSE, col = c("yellow","beige"),main="ALASKA / AM West: Delayed Ratio Median", names.arg=c("ALASKA","AM WEST"), xlab = "Airlines")
```
Creating the **TotFlights**,**DelayedRatio**, and **OnTimeRatio** allows us to explore deeper and see how the the airlines have a close median of on time and delayed flights.









